Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-09-02 summary.Rmd
typos

We will use the notation $A \sim PABM(\{\lambda^{(kl)}\}_K)$ to denote a random 
adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
consisting of $K$ underlying clusters/communities.

@@

1. Cluster membership identification (up to permutation).
2. Parameter estimation (estimating $\lambda^{(kl)}$'s).

Here, we will focus more on (1) and pose possible methods for (2). In our 
methods, we assume that $K$, the number of clusters, is known beforehand and 
does not require estimation.

@@

\citet*{noroozi2019estimation} proposed using sparse subspace clustering (SSC) 
to identify the cluster memberships given either an edge probability matrix $P$ 
or an adjacency matrix $A$. In the case that $P$ is known, the cluster 
memberships can be identified exactly (up to permutation). A similar procedure 
can be applied if $P$ is unknown and we have an observation $A$, but the 
theoretical guarantees of this method applied to the PABM are unknown. In 
particular, the method requires spherical Gaussian noise. The authors of this 
paper then use point estimators for $\{\lambda^{(kl)}\}$ using cluster labels 
obtained via SSC.

## Clustering

@@

We previously stated one possible set of latent positions that result in the 
edge probability matrix of a PABM, $P = (XU) I_{pq} (XU)^\top$. If we have 
(or can estimate) $XU$ directly, then both the clustering and parameter 
identification problem are trivial since $U$ is orthonormal and fixed for each 
value of $K$. However, direct identification or estimation of $XU$ is not 
possible \cite{rubindelanchy2017statistical}.

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the RDPG, 
$Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different clusters.

We can note from the explicit form of $XU$ that it represents points in 
$\mathbb{R}^{K^2}$ such that points within each cluster lie on $K$-dimensional 
subspaces. Furthermore, the subspaces are orthogonal to each other. 
Multiplication by $Q \in \mathbb{O}(p, q)$ removes the orthogonality property 
but retains the property that each cluster is represented by a $K$-dimensional 
subspace. Using this property, previous work proposes the use of subspace 
clustering while acknowledging some of its shortcomings 
\cite{noroozi2019estimation} \cite{soltanolkotabi2014}.

**Theorem 3.** 
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix. Let $B = V V^\top$. Then $B_{ij} = 0$ if vertices $i$ and $j$ are 
of different clusters.

*Proof (sketch)*
By projection, $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is defined as 
in Theorem 2. Since $X$ is block diagonal with each block corresponding 
to one cluster, $X (X^\top X)^{-1} X^\top$ is also a block diagonal matrix 
with each block corresponding to a cluster and zeros elsewhere. Therefore, 
if vertices $i$ and $j$ belong to different clusters, then the $ij$^th^ element 
of $X (X^\top X)^{-1} X^\top = V V^\top = B$ is 0.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Edge probability matrix $P$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the spectral decomposition $P = V D V^\top$.\;
    Compute the inner product matrix $B = |V V^\top|$, applying $|\cdot|$ 
    entry-wise.\;
    Construct graph $G$ using $B$ as edge similarities.\;
    Identify the connected components of $G$ and map each to cluster labels 
    $1, ..., K.$\;
  \caption{PABM clustering on the edge probability matrix.}
\end{algorithm}

@@

The adjacency embedding of $A$ approaches latent positions that form $P$ as 
the number of vertices $n$ increases. More 
precisely, let $\{\lambda^{(kl)}\}_K \sim \mathcal{F}_K$ for some joint 
distribution consisting of $K$ underlying clusters $\mathcal{F}_K$. Then the 
latent positions $XU \sim \mathcal{G}_K$ for some related joint distribution 
with $K$ underlying clusters $\mathcal{G}_K$. Denote $Z_n$ as a sample of size 
$n$ from $\mathcal{G}_K$ and adjacency matrix $A_n$ as one draw from edge 
probability matrix $P_n = Z_n I_{pq} Z_n^\top$. Let $\hat{Z}_n$ be the 
adjacency embedding of $A_n$ with rows $(\hat{z}_i^{(n)})^\top$. Then by
\citet*{rubindelanchy2017statistical},

@@

**Theorem 4**. Let $\hat{V}^{(n)} \in \mathbb{R}^{n \times K^2}$ be the matrix 
of $K^2$ eigenvectors of $A_n$ corresponding to the $K (K+1) / 2$ most positive 
eigenvalues and $K (K-1) / 2$ most negative eigenvalues with rows
$(\hat{v}_i^{(n)})^\top$. Let $(i, j)$ correspond to pairs belonging to 
different clusters. Then for some $c > 0$,

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected components 
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the cluster labels $1, ..., K$.\;
  \caption{PABM clustering on the adjacency matrix.}
\end{algorithm}

@@

For any $P$ edge probability matrix for the PABM such that the rows and 
columns are organized by cluster, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda%{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P = \sigma^2 u v^\top$ be the 
singular value decomposition of $P$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the edge probability matrix.}
  \KwData{Edge probability matrix $P$, cluster assignments $1, ..., K$.}
  \KwResult{PABM parameters $\{\lambda^{(kl)}\}_K$}
  Arrange the rows and columns of $P$ by cluster such that each $P^{(kl)}$ 
  block consists of edge probabilities between clusters $k$ and $l$.\; 
  \For{$k, l = 1, ..., K$, $k \geq l$} {
    Compute $P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top$, the SVD of 
    the $kl^{th}$ block.\;
    Assign $\lambda^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\lambda^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the adjacency matrix.}
  \KwData{Adjacency matrix $A$, cluster assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Arrange the rows and columns of $A$ by cluster such that each 
  $A^{(kl)}$ block consists of estimated edge probabilities between 
  clusters $k$ and $l$.\;
  \For{$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $\sigma^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

@@

*Proof (sketch)*. 
Let $P$ and $A$ be organized by community such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to the edges between clusters $k$ and $l$.

@@

* Number of underlying clusters $K = 2$
* Mixture parameter $\alpha = .5$
* Within-cluster popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-cluster popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

### Clustering

In this part, we will assess Algorithm 2's performance for sample sizes 
$n = 64, 128, 256, 512, 1024, 2048$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered. We will not consider clustering using the 
edge probability matrix $P$ since this will always result exact recovery of the
original clusters (up to permutation).

@@

* Number of underlying clusters $K = 2, 4, 8$
* Mixture parameters $\alpha_k = 1 / K$ for $k = 1, ..., K$
* Within-cluster popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-cluster popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

### Clustering

>>>>

We will use the notation $A \sim PABM(\{\lambda^{(kl)}\}_K)$ to denote a random 
adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
consisting of $K$ underlying communities.

@@

1. Community membership identification (up to permutation).
2. Parameter estimation (estimating $\lambda^{(kl)}$'s).

Here, we will focus more on (1) and pose possible methods for (2). In our 
methods, we assume that $K$, the number of communities, is known beforehand and 
does not require estimation.

@@

\citet*{noroozi2019estimation} proposed using sparse subspace clustering (SSC) 
to identify the community memberships given either an edge probability matrix $P$ 
or an adjacency matrix $A$. In the case that $P$ is known, the community 
memberships can be identified exactly (up to permutation). A similar procedure 
can be applied if $P$ is unknown and we have an observation $A$, but the 
theoretical guarantees of this method applied to the PABM are unknown. In 
particular, the method requires spherical Gaussian noise. The authors of this 
paper then use point estimators for $\{\lambda^{(kl)}\}$ using cluster labels 
obtained via SSC.

## Community detection

@@

We previously stated one possible set of latent positions that result in the 
edge probability matrix of a PABM, $P = (XU) I_{pq} (XU)^\top$. If we have 
(or can estimate) $XU$ directly, then both the community detection and parameter 
identification problem are trivial since $U$ is orthonormal and fixed for each 
value of $K$. However, direct identification or estimation of $XU$ is not 
possible \cite{rubindelanchy2017statistical}.

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the RDPG, 
$Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different communities.

We can note from the explicit form of $XU$ that it represents points in 
$\mathbb{R}^{K^2}$ such that points within each community lie on $K$-dimensional 
subspaces. Furthermore, the subspaces are orthogonal to each other. 
Multiplication by $Q \in \mathbb{O}(p, q)$ removes the orthogonality property 
but retains the property that each community is represented by a $K$-dimensional 
subspace. Using this property, previous work proposes the use of subspace 
clustering while acknowledging some of its shortcomings 
\cite{noroozi2019estimation} \cite{soltanolkotabi2014}.

**Theorem 3.** 
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix. Let $B = V V^\top$. Then $B_{ij} = 0$ if vertices $i$ and $j$ are 
of different communities.

*Proof (sketch)*
By projection, $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is defined as 
in Theorem 2. Since $X$ is block diagonal with each block corresponding 
to one community, $X (X^\top X)^{-1} X^\top$ is also a block diagonal matrix 
with each block corresponding to a community and zeros elsewhere. Therefore, 
if vertices $i$ and $j$ belong to different communities, then the $ij$^th^ 
element of $X (X^\top X)^{-1} X^\top = V V^\top = B$ is 0.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Edge probability matrix $P$, number of communities $K$}
  \KwResult{Community labels $1, ..., K$}
    Compute the spectral decomposition $P = V D V^\top$.\;
    Compute the inner product matrix $B = |V V^\top|$, applying $|\cdot|$ 
    entry-wise.\;
    Construct graph $G$ using $B$ as edge similarities.\;
    Identify the connected components of $G$ and map each to community labels 
    $1, ..., K.$\;
  \caption{PABM clustering on the edge probability matrix.}
\end{algorithm}

@@

The adjacency embedding of $A$ approaches latent positions that form $P$ as 
the number of vertices $n$ increases. More 
precisely, let $\{\lambda^{(kl)}\}_K \sim \mathcal{F}_K$ for some joint 
distribution consisting of $K$ underlying communities $\mathcal{F}_K$. Then the 
latent positions $XU \sim \mathcal{G}_K$ for some related joint distribution 
with $K$ underlying communities $\mathcal{G}_K$. Denote $Z_n$ as a sample of size 
$n$ from $\mathcal{G}_K$ and adjacency matrix $A_n$ as one draw from edge 
probability matrix $P_n = Z_n I_{pq} Z_n^\top$. Let $\hat{Z}_n$ be the 
adjacency embedding of $A_n$ with rows $(\hat{z}_i^{(n)})^\top$. Then by
\citet*{rubindelanchy2017statistical},

@@

**Theorem 4**. Let $\hat{V}^{(n)} \in \mathbb{R}^{n \times K^2}$ be the matrix 
of $K^2$ eigenvectors of $A_n$ corresponding to the $K (K+1) / 2$ most positive 
eigenvalues and $K (K-1) / 2$ most negative eigenvalues with rows
$(\hat{v}_i^{(n)})^\top$. Let $(i, j)$ correspond to pairs belonging to 
different communities. Then for some $c > 0$,

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected components 
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{PABM clustering on the adjacency matrix.}
\end{algorithm}

@@

For any $P$ edge probability matrix for the PABM such that the rows and 
columns are organized by community, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda%{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P = \sigma^2 u v^\top$ be the 
singular value decomposition of $P$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the edge probability matrix.}
  \KwData{Edge probability matrix $P$, community assignments $1, ..., K$.}
  \KwResult{PABM parameters $\{\lambda^{(kl)}\}_K$}
  Arrange the rows and columns of $P$ by community such that each $P^{(kl)}$ 
  block consists of edge probabilities between communities $k$ and $l$.\; 
  \For{$k, l = 1, ..., K$, $k \geq l$} {
    Compute $P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top$, the SVD of 
    the $kl^{th}$ block.\;
    Assign $\lambda^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\lambda^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the adjacency matrix.}
  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Arrange the rows and columns of $A$ by community such that each 
  $A^{(kl)}$ block consists of estimated edge probabilities between 
  communities $k$ and $l$.\;
  \For{$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $\sigma^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

@@

*Proof (sketch)*. 
Let $P$ and $A$ be organized by community such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to the edges between communities $k$ and 
$l$.

@@

* Number of underlying communities $K = 2$
* Mixture parameter $\alpha = .5$
* Within-group popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

### Community detection

In this part, we will assess Algorithm 2's performance for sample sizes 
$n = 64, 128, 256, 512, 1024, 2048$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered. We will not consider community detection using the 
edge probability matrix $P$ since this will always result exact recovery of the
original communities (up to permutation).

@@

* Number of underlying communities $K = 2, 4, 8$
* Mixture parameters $\alpha_k = 1 / K$ for $k = 1, ..., K$
* Within-group popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

### Community detection
